---
title: "thesis analysis"
output:
  html_document: default
subtitle: 
author: mandy osuji
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
# install.packages("arm")
# install.packages("blme")
# install.packages("psycho")
# install.packages("glmnet")
# install.packages("car")
# install.packages("multcomp")
install.packages("ggthemes")
library(arm)
library(lme4)
library(blme)
library(ggplot2)
library(stringr)
library(psycho)
library(glmnet)
library(car)
library("multcomp")
library(ggthemes)
```

#### First, load the results data from output.csv and take a peek.

```{r}
# load csv file
results <- read.csv(file = 'data/output.csv', header = TRUE)
print(head(results), row.names = FALSE)
```

#### Added columns for the log reaction time and that value centered on each participant's mean, which will be helpful for later analysis and visualization.

```{r}
# add colums for log(rt) and mean-centered log(rt)
results$rt_log <- log(results$rt)
results$rt_log_centered <- results$rt_log - ave(results$rt_log, results$id, FUN = function(x) mean(x, na.rm = TRUE))
print(results[1:10, c("n1", "n2", "rt", "rt_log", "rt_log_centered")], row.names = FALSE)
#exclude practice trials
results <- results[results$relation_type != "practice", ]
```

#### Response time by relation

```{r}
# box and whisker plots

# centered log rt, split by relation type
pdf("graphs/rt-by-relation.pdf", width=6, height=4)
ggplot(results, aes(x = relation_type, y = rt_log_centered)) + 
  geom_boxplot(size=.25, fill="thistle", staplewidth = 0.5, outlier.shape = 1, outlier.size = 1) + 
  labs(x = "Relation type",
       y = "Centered log response time") +
  theme_bw() +
  theme(panel.grid = element_blank(), axis.title.x = element_text(margin = margin(t = 10)))
dev.off()

# box and whisker plot for all participants
pdf("graphs/rt-by-participant.pdf", width=6, height=4)
ggplot(results, aes(x = factor(id), y = rt_log)) + 
  geom_boxplot(size=.25, fill="lightgoldenrod", staplewidth = 0.5, outlier.shape = 1, outlier.size = 1) + 
  labs(x = "Participant",
       y = "Log response time") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 6), panel.grid = element_blank(), axis.title.x = element_text(margin = margin(t = 10)))
dev.off()

# pdf("graphs/rt-by-relation-hist.pdf", width=6, height=4)
# ggplot(results, aes(x = rt_log_centered, fill=relation_type)) + 
#   geom_histogram(bins=50, position = "identity", alpha = 0.5) + 
#   labs(x = "Centered log response time",
#        y = "",
#        fill="Relation type") +
#   theme_bw() +
#   theme(legend.position="bottom")
# dev.off()
```

#### Scatter plots

```{r}
# log rt
# ggplot(results, aes(x=n_sim, y=rt_log)) + 
#   ggtitle("GloVe Similarity vs Log Response Time") +
#   geom_jitter(size=1.2, alpha=0.5, color="darkorchid2")+ 
#   geom_smooth(method=lm , formula = y ~ x, color="tomato", se=TRUE) +
#   labs(x = "GloVe similarity", y = "Log response time")+
#   theme_bw()

# centered log rt
# ggplot(results, aes(x=n_sim, y=rt_log_centered)) + 
#   ggtitle("GloVe Similarity vs Centered Log Response Time") +
#   geom_jitter(width = 0.01, height=0, size=1.2, alpha=0.5, color="cornflowerblue") + 
#   geom_smooth(method=lm , formula = y ~ x, color="tomato", se=TRUE) +
#   labs(x = "GloVe similarity", y = "Centered log response time")+
#   theme_bw()

# this one colors according to different relations
# ggplot(results, aes(x=n_sim, y=rt_log, color=factor(relation_type))) + 
#   ggtitle("GloVe Similarity vs Log Response Time") +
#   geom_point(size=1.2) + 
#   labs(x = "GloVe similarity", y = "Log response time", color="Relation type")+
#   theme_bw()

# scatterplot with x-axis glove similarity and y-axis rt
# glove similarity versus centered log response time
pdf("graphs/rt-glove-scatter.pdf", width=6, height=4)
ggplot(results, aes(x=n_sim, y=rt_log_centered, color=factor(relation_type))) + 
  geom_jitter(width = 0.005, height=0, size=1.2, alpha=0.5) +
  labs(x = "GloVe similarity", y = "Centered log response time", color="Relation type") +
  geom_smooth(method=lm , formula = y ~ x, color="darkmagenta", linewidth=.75, se=TRUE) +
  theme_bw() +
  theme(legend.position="bottom", axis.title.x = element_text(margin = margin(t = 20)))
dev.off()
```

#### Some summary stats

```{r}
# summary stats of raw response times for each relation type
summary(results$rt[results$relation_type == "hypernym"])
summary(results$rt[results$relation_type == "meronym"])
summary(results$rt[results$relation_type == "sister"])
summary(results$rt[results$relation_type == "synonym"])
```


#### Linear Mixed-Effects Models

```{r}
# first we need to use sum-coding on relation_type since it is categorical
relation_types <- c("sister", "hypernym", "meronym", "synonym")
results$relation_type = factor(results$relation_type, levels=relation_types)
contrasts(results$relation_type) <- contr.sum(4)
# results
```

```{r}
# now we can make our lme model to predict reaction time based on glove similarity, relation type, and response length, controlling for differences between individual participants
relations.model <- lmer(rt ~ n_sim + relation_type + response_length + (1|id), data=results)
summary(relations.model)
```


```{r}
# in order to test for significance, we will do a likelihood ratio test between the full model and a null model without the factor of interest
relations.model <- lmer(rt ~ n_sim + relation_type + (1|response_length) + (1|id), data=results, REML=FALSE)
summary(relations.model)

# here the factor of interest is glove similarity, and we do see significance
relations.model.null <- lmer(rt ~ relation_type + (1|response_length) + (1|id), data=results, REML=FALSE)
anova(relations.model.null, relations.model)

# here the factor of interest is relation type, and we see no significance
relations.model.null <- lmer(rt ~ n_sim + (1|response_length) + (1|id), data=results, REML=FALSE)
anova(relations.model.null, relations.model)
# if relation type was significant we would do the following to get more specifics on category influence:
# relations_model_multcomp <- glht(relations.model.with.types, linfct=mcp(relation_type="Tukey"))
# summary(relations_model_multcomp)

# multicollinarity check, all VIF close to 1 indicating multicollinarity not a concern for interpretation
vif(relations.model)
```

```{r}
# # old initial models without relation_type
# 
# #relations.model <- glmer(rt ~ n_sim + response_length + (1|id), data=results)
# relations.model <- lmer(rt ~ n_sim + response_length + (1|id), data=results)
# summary(relations.model)
# 
# # likelihood ratio test, factor of interest is GloVe similarity
# relations.null <- lmer(rt ~ response_length + (1|id), data=results, REML=FALSE)
# relations.model <- lmer(rt ~ n_sim + response_length + (1|id), data=results, REML=FALSE)
# # coef(relations.model)
# anova(relations.null, relations.model)
# 
# # likelihood ratio test, factor of interest is response length
# relations.null <- lmer(rt ~ (1|id), data=results, REML=FALSE)
# relations.model <- lmer(rt ~ response_length + (1|id), data=results, REML=FALSE)
# # coef(relations.model)
# anova(relations.null, relations.model)
```

#### Visualizing cosine similarity between responses and base nouns?

```{r}
similarities <- read.csv(file = 'data/sims.csv', header = TRUE)
similarities <- similarities[similarities$relation_type != "practice", ]
print(head(similarities), row.names = FALSE)
```

```{r}
# see how GloVe similarity is distributed among relation types
pdf("graphs/glove-by-relation.pdf", width=6, height=4)
ggplot(similarities, aes(x = relation_type, y = n_sim)) + 
  geom_boxplot(size=.5, fill="lemonchiffon2") + 
  labs(x = "Relation type",
       y = "GloVe similarity between nouns in pair") +
  theme_bw()
dev.off()
```

```{r}
# ggplot(subset(similarities, !is.na(avg_sim)), aes(x = r_n1_sim, y = r_n2_sim)) +
#   geom_point(alpha = 0.5, color = "steelblue") +
#   geom_abline(slope = 1, intercept = 0, linetype = "solid", color = "tomato") +
#   labs(
#     title = "similarity of response to n1 vs n2",
#     x = "Similarity to n1",
#     y = "Similarity to n2"
#   ) + 
#   xlim(0, 1) + ylim(0, 1) +
#   theme_bw()
```
```{r}
# with_lin <- read.csv(file = 'data/extra.csv', header = TRUE)
# with_lin <- with_lin[with_lin$relation_type != "practice", ]
# print(head(with_lin), row.names = FALSE)
```


```{r}
# try model including lin similarity
# relations.null <- lmer(rt ~ n_sim + response_length + (1|id), data=with_lin, REML=FALSE)
# relations.model <- lmer(rt ~ n_sim + lin_similarity + response_length + (1|id), data=with_lin, REML=FALSE)
# anova(relations.null, relations.model)
```
```{r}
# GLOVE VISUAL (this takes awhile...)
#install.packages("Rtsne")
library(Rtsne)

# read in data from file, format
glove <- read.csv("data/glove.300d.txt", sep = " ", header = FALSE, quote = "")
glove_subset <- glove[1:10000, ]
word_labels <- glove_subset[, 1]
vectors_subset <- as.matrix(glove_subset[, -1])
rownames(vectors_subset) <- word_labels

# do dimension reduction
set.seed(42)
tsne_out <- Rtsne(vectors_subset, dims = 2, perplexity = 30)

# words to emphasize
target_words <- c("boat", "ocean", "trip", "dock", 
                  "thesis", "learning", "work", "student",
                  "dress", "style", "scarf", "necklace")

# create data frame for plotting
plot_data <- data.frame(
  x = tsne_out$Y[, 1],
  y = tsne_out$Y[, 2],
  word = word_labels
)
plot_data$label <- ifelse(plot_data$word %in% target_words, as.character(plot_data$word), "")

# generate the scatter plot
pdf("graphs/glove-cloud.pdf", width=6, height=4)
ggplot(plot_data, aes(x, y)) +
  geom_point(alpha = 0.2, color = "grey") + # Faded background points
  geom_text(aes(label = label), color = "blue4", fontface = "bold", size = 3) + 
  theme_minimal() +
  labs(x = "", y="") +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
dev.off()
```

